---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sleepcycles: Detect and Visualize Sleep Cycles from Hypnograms

<!-- badges: start -->

[![R-CMD-check](https://github.com/JasonDude16/sleepcycles/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JasonDude16/sleepcycles/actions/workflows/R-CMD-check.yaml) ![GitHub release](https://img.shields.io/github/v/release/JasonDude16/sleepcycles) [![Codecov test coverage](https://codecov.io/gh/JasonDude16/sleepcycles/graph/badge.svg)](https://app.codecov.io/gh/JasonDude16/sleepcycles) [![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)

<!-- badges: end -->

The `sleepcycles` package provides tools for detecting, analyzing, and visualizing sleep cycles from hypnogram data using density-based and rule-based (Feinberg) algorithms. It supports both single- and multi-subject hypnograms, enabling users to extract and explore Non-REM (NREMP) and REM (REMP) periods with flexible parameter settings. The package includes an interactive Shiny application, a suite of visualization functions and utilities validating sleep staging data.

📌 Designed for sleep researchers, clinicians, and data scientists, `sleepcycles` helps uncover the underlying dynamics of sleep architecture and assess the impact of sleep patterns in healthy and clinical populations.

------------------------------------------------------------------------

## **📦 Installation**

You can install the development version from GitHub using `{remotes}`:

``` r
# Install remotes if not installed
install.packages("remotes")

# Install sleepcycles from GitHub
remotes::install_github("jasondude16/sleepcycles")
```

## 🚀 Quick Start Guide

#### 1️⃣ Load the Package

```{r}
library(sleepcycles)
```

#### 2️⃣ Load Example Data

`sleepcycles` contains two example datasets: `hypnogram_single` and `hypnogram_grouped`, which contain single- and multi-subject hypnograms, respectively. We will load the multi-subject hypnogram for this example.

```{r}
data("hypnogram_grouped", package = "sleepcycles")
```

`sleepcycles` requires a hypnogram (in the format of a data frame) with two variables: one of the epoch, and one of the sleep stage; if there are multiple subjects, an ID variable is also required.

```{r}
head(hypnogram_grouped)
```

#### 3️⃣ Detect Sleep Cycles

```{r warning=FALSE}
sleepcycle_obj <- sleepcycles_from_hypnogram(
  hypnogram_grouped, 
  epoch_col = "epoch", 
  stage_col = "stage", 
  id_col = "id",
  verbose = FALSE
)
```

#### 4️⃣ Visualize Sleep Cycles

```{r fig.height=3}
plot_cycles(sleepcycle_obj, id = "1")
```

```{r fig.height=3}
plot_hypnogram(sleepcycle_obj, id = "1")
```

#### 5️⃣ Run the Interactive Shiny App

```{r}
if (interactive()) {
  run_sleepcycles_app(sleepcycle_obj)
}
```

### 📊 Features

-   ✔ **Sleep Cycle Detection**: Supports Feinberg (Rule-Based) algorithm and a novel density-based algorithm
-   ✔ **Interactive Visualization**: Generate hypnograms, density plots, and cycle summaries
-   ✔ **Shiny App**: Modify and explore cycle detection parameters interactively
-   ✔ **Preprocessing & Validation**: Built-in data validation tools
-   ✔ **Single and Grouped Data Support**: Works with individual and multi-subject hypnogram datasets

### 📖 Available Functions

| **Function**                   | **Description**                                                 |
|----------------------------|--------------------------------------------|
| `sleepcycles_from_hypnogram()` | Detects sleep cycles using DUDE or Feinberg method              |
| `plot_hypnogram()`             | Generates a hypnogram plot                                      |
| `plot_densities()`             | Plots hypnogram stage densities                                 |
| `plot_cycles()`                | Visualizes detected sleep cycles                                |
| `plot_summary()`               | Provides a combined visualization of sleep cycles and hypnogram |
| `run_sleepcycles_app()`        | An interactive Shiny app for sleep cycle parameter exploration  |
| `check_hypnogram()`            | Validates hypnogram data for errors                             |

### 📌 Contributing

We welcome contributions! If you find a bug, please [open an issue](https://github.com/JasonDude16/sleepcycles/issues).

### 📜 License

This package is licensed under the MIT License.
