---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# sleepcycles: Detect and Visualize Sleep Cycles from Hypnograms

<!-- badges: start -->

<!-- badges: end -->

**Analyze, Detect, and Visualize Sleep Cycles with Ease.**

The **sleepcycles** package provides tools for detecting, analyzing, and visualizing **sleep cycles** from hypnogram data using **density-based (DUDE) and rule-based (Feinberg) algorithms**. It supports both **single-subject** and **grouped** datasets, enabling users to extract and explore **Non-REM (NREMP) and REM (REMP) periods** with flexible parameter settings. The package includes **interactive Shiny applications**, a suite of **visualization functions** (e.g., hypnograms, density plots, and cycle summaries), and utilities for **preprocessing and validating sleep staging data**.

📌 **Designed for sleep researchers, clinicians, and data scientists, `sleepcycles` helps uncover the underlying dynamics of sleep architecture and assess the impact of sleep patterns on health and performance.**

------------------------------------------------------------------------

## **📦 Installation**

You can install the development version from GitHub using `{remotes}`:

``` r
# Install remotes if not installed
install.packages("remotes")

# Install sleepcycles from GitHub
remotes::install_github("yourusername/sleepcycles")
```

🚀 Quick Start Guide

1️⃣ Load the Package

```{r}
library(sleepcycles)
```

2️⃣ Load Example Data

```{r}
data("hypnogram_single", package = "sleepcycles")

# Preview the dataset
head(hypnogram_single)
```

3️⃣ Detect Sleep Cycles

```{r}
sleepcycles_obj <- sleepcycles_from_hypnogram(
  hypnogram_single, 
  epoch_col = "epoch", 
  stage_col = "stage", 
  method = "dude"
)
```

4️⃣ Visualize Sleep Cycles

```{r}
plot_summary(sleepcycles_obj)
```

5️⃣ Run the Interactive Shiny App

```{r}
if (interactive()) {
  run_sleepcycles_app(sleepcycles_obj)
}
```

📊 Features
✔ Sleep Cycle Detection: Supports DUDE (Density-Based) and Feinberg (Rule-Based) algorithms
✔ Interactive Visualization: Generate hypnograms, density plots, and cycle summaries
✔ Shiny App: Modify and explore cycle detection parameters interactively
✔ Preprocessing & Validation: Built-in data validation tools
✔ Single and Grouped Data Support: Works with individual and multi-subject datasets

📖 Available Functions

| **Function**                      | **Description**                                         |
|------------------------------------|---------------------------------------------------------|
| `sleepcycles_from_hypnogram()`     | Detects sleep cycles using DUDE or Feinberg            |
| `plot_hypnogram()`                 | Generates a hypnogram plot                             |
| `plot_densities()`                 | Plots sleep cycle density curves                      |
| `plot_cycles()`                    | Visualizes detected sleep cycles                      |
| `plot_summary()`                   | Provides a combined visualization of sleep cycles     |
| `run_sleepcycles_app()`            | Launches an interactive Shiny app for cycle exploration |
| `check_hypnogram()`                | Validates hypnogram data for errors                   |

📚 Example Datasets
The package includes two example datasets:

| **Dataset**           | **Description**                                    |
|----------------------|----------------------------------------------------|
| `hypnogram_single`  | Example hypnogram data for a single individual     |
| `hypnogram_grouped` | Example hypnogram data for multiple individuals    |

```{r}
data("hypnogram_single", package = "sleepcycles")
data("hypnogram_grouped", package = "sleepcycles")
```

📌 Contributing
We welcome contributions! If you find a bug, please [open an issue](https://github.com/JasonDude16/sleepcycles/issues).

📜 License
This package is licensed under the MIT License.

📞 Contact
📧 Jason Dude
🔗 [GitHub][https://github.com/JasonDude16]
🐦 Your Twitter Handle

















